---
title: "Explore_Loan_Data"
author: "Brahmi Ibtissam"
date: "9 septembre 2018"
output: html_document
fig_width: 12 
fig_height: 4 
editor_options: 
  chunk_output_type: console
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
#Load the data
df=read.csv('prosperLoanData.csv')
```
> We can see the variables descriptions of the data at the link
> https://docs.google.com/spreadsheets/d/1gDyi_L4UvIrLTEC6Wri5nbaMmkGmLQBk-Yx3z0XDEtI/edit#gid=0

>To help ourselves understand the variables and the relationship between them and answer relevant questions, and since we have 81 variables, we'll select the key ones by subseting our dataset.
> We'll choose  
>LoanOriginalAmount
>BorrowerRate
>LoanStatus
>IncomeRange
>BorrowerState
>ListingCategory
>EmploymentStatus
>Occupation
>CurrentCreditLines
>ListingKey
>LoanKey
>Term
>OnTimeProsperPayments.

##Subsetting the data
```{r echo=FALSE, message=FALSE, warning=FALSE}
#Select more relevant variables from the whole dataset (df) and store them in df_s
library(dplyr)
df_s <- df %>%
  select(LoanOriginalAmount, BorrowerRate, LoanStatus, IncomeRange, BorrowerState,Occupation,ListingCategory..numeric., Term, OnTimeProsperPayments , EmploymentStatus, CurrentCreditLines, ListingKey, LoanKey,Term)
```
#Univariate Analysis
##First to get realy started with our data and get deep understanding of it, we'll produce summaries and visualizations of each individual variable
```{r echo=FALSE, message=FALSE, warning=FALSE}
#get the dimension of our whole dataset
dim(df)
```
> Our Dataset consists of 81 variables with almost 113937 observations, we reduced the number of variable to 13.


```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
#Plot the distribution of loan original amount 
require(ggplot2)
ggplot(data=df_s, aes(x=LoanOriginalAmount)) +
  geom_histogram(binwidth=500, color= I('black'), fill=I('#099DD9')) +
  scale_x_continuous(breaks=seq(1000,30000,1000), limits=c(1000,30000))
```

We notice the presence of 3 peaks: about 15000 borrowers borrow an amount of 4000 dollars, 10000 dollars is  borrowed by about 12000 of Prosper customers, and finally the third peak indicates that 12500 of customers have an original amount loan of 15000 dollars, maybe the three peaks are explained by the difference in monthly incomes and needs.

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
#Plot the distribution of on time prosper payments
ggplot(data=df_s, aes(x=OnTimeProsperPayments)) +
  geom_histogram( color= I('black'), fill=I('#099DD9'))+
  scale_x_continuous(breaks=seq(0,100,5), limits=c(0,100))
```

We notice that more than 5000 of borrowers pay on time just 10 times.
The distribution is left skewed with a long tail.



```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
#Distribution of current credit lines
ggplot(data=df_s, aes(x=CurrentCreditLines)) +
  geom_histogram( color= I('black'), fill=I('#099DD9')) +
  scale_x_continuous(breaks=seq(0,40,2), limits=c(0,40))+
  scale_y_continuous(breaks=seq(0,20000,1000), limits=c(0,20000))
```

 More than 16000 of borrowers have 10 credit lines at the time the credit profile was pulled.


```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
#Plot the distribution of loan original amounts of each state separately
ggplot(data=df_s, aes(x=LoanOriginalAmount)) +
  geom_histogram(binwidth=500, color= I('black'), fill=I('#099DD9')) +
  scale_x_continuous(breaks=seq(1000,30000,1000), limits=c(1000,30000))+
  facet_wrap(~BorrowerState)
```

Most of borrowers are from big cities where the standard of living is somewhat expensive: California New York Texas Florida...

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
#Transform Listing category to a factor variable to facet with
df_s$ListingCategory <- factor(df_s$ListingCategory)
str(df_s$ListingCategory)
```

 We transformed ListingCategory to a factor variable such that we can facet with it

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
#Plot the distribution of loan original amount for each listing category 
ggplot(data=df_s, aes(x=LoanOriginalAmount)) +
  geom_histogram(binwidth=500, color= I('black'), fill=I('#099DD9')) +
    facet_wrap(~ListingCategory)
```

Most of demanded loans are about debt consolidation  
```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
#get the structure of income range variable
str(df_s$IncomeRange)
```

We arrange our data set by Income Range

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
#arrange by income range to have a well presented plot
library(dplyr)
df_s <- df_s %>%
 arrange(IncomeRange)
```

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #plot the box plot of income range Vs loan original amount
ggplot(data=subset(df_s, !is.na(IncomeRange)), aes(y=LoanOriginalAmount, x=IncomeRange)) +
  geom_boxplot()
```

We notice the presence of outliers, especially for 1$-24,900$ class, borrowers with high incomes have the greatest IQR and median.

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #plot the box plot of occupation Vs loan original amount
ggplot(data=subset(df_s, !is.na(IncomeRange)), aes(y=LoanOriginalAmount, x=Occupation)) +
  geom_boxplot()
```

It's interesting to see whats high loan original amount borrowers occupation, but we have so much levels that we can't visualize in a clean manner.


#Bivariate Analysis
```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #plot the scatter plot of loan original amount and borrower rate
ggplot(data=subset(df_s, !is.na(IncomeRange)), aes(x=LoanOriginalAmount, y=BorrowerRate)) +
  geom_point(alpha=0.7, position='jitter')
```

The log scale will be more informative since we have a multiplicative relationship between x and y
```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #reproduce the same preceding plot with a log scale of x-axis
ggplot(data=subset(df_s, !is.na(IncomeRange)), aes(x=LoanOriginalAmount, y=BorrowerRate)) +
  geom_point(alpha=0.5, size=0.5, position='jitter')+
  scale_x_log10()
```

Besides over plotting, we see that for the same loan original amount we have a lot of degrees of borrower rates?!

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #A plot showing the correlation between the current credit lines and on time prosper payments
ggplot(data=subset(df_s, !is.na(CurrentCreditLines)), aes(x=CurrentCreditLines, y=OnTimeProsperPayments)) +
  geom_point(alpha=0.5, size=0.5, position='jitter')
```


```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #calculate the Pearson correlation between current credit lines and on time prosper payments
cor.test(df_s$CurrentCreditLines,df_s$OnTimeProsperPayments)
```

Expecting that it will be a negative correlation between current credit lines and on time prosper payments, it turns to be 0.06! the distribution seems to be somewhat normal and we also notice the presence of some outliers; for example, someone has about 60 credit lines at a time, but not even 50 on time payments!

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #plot the correlation between listing category and on time prosper payments
ggplot(data=subset(df, !is.na(IncomeRange)), aes(x=ListingCategory..numeric., y=OnTimeProsperPayments)) +
  geom_point()+
  scale_x_continuous(breaks=seq(0,20,1))+
  scale_y_continuous(breaks=seq(0,150,10))
```

Most of on time prosper payments concern debt consolidation and home improvement, comes in second range Household Expenses and Large Purchases.

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #Scatter plot between Term (in months) and borrower rate
ggplot(data=subset(df, !is.na(IncomeRange)), aes(x=Term, y=BorrowerRate)) +
  geom_point( size=0.5, position='jitter')
```

Borrowers who choose 36 months as a length of the loan (middle term) have the highest Borrower Rate

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #Scatter plot between Term (in months) and on time prosper payments
ggplot(data=subset(df, !is.na(IncomeRange)), aes(x=Term, y=OnTimeProsperPayments)) +
  geom_point( size=0.5, position='jitter')
```

On time Prosper payments is frequent for borrowers who choose the middle term, maybe we can ask who what is the Income range of those borrowers!

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
  #Scatter plot between income range and on time prosper payments for avery Term (in months) 
ggplot(data=subset(df, !is.na(IncomeRange)), aes(x=IncomeRange, y=OnTimeProsperPayments)) +
  geom_point( size=0.5, position='jitter')+
  facet_wrap(~Term,ncol=1)
```

Borrowers with in income range of 25000 and 74999 dollars and divided their loan on 36 months are more prone to pay on time.

Now we'll be interested to the mean of original amount loan by income range, before, we should order Income Range factor variable

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #arrange the factors variable income range from lowest to highest
df_s$IncomeRange<-ordered(df$IncomeRange, level=c('$0', '$1-24,999', '$25,000-49,999', '$50,000-74,999', '$75,000-99,999', '$100,000+' , 'Not displayed', 'Not employed'))
```

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #plot the income range Vs the mean loan original amount
ggplot(subset(df, !is.na(IncomeRange)), aes(IncomeRange, LoanOriginalAmount, group=1))+
  geom_line(stat='summary', fun.y=mean)
```

Except the income range between 0$ and 24,999$, as we climb in income ranges the loan original amount increases.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#create a reduced dataset containing less variables to plot a scatter plot matrix
library(dplyr)
df_s_gg <- df %>%
  select(LoanOriginalAmount, BorrowerRate, LoanStatus, IncomeRange,ListingCategory..numeric., Term, OnTimeProsperPayments , EmploymentStatus, CurrentCreditLines)
```


```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #plot a scatter plot matrix
library(GGally)
set.seed(1598)
theme_set(theme_minimal(20))
ggpairs(df_s_gg[sample.int(nrow(df_s_gg),1000),])
```

We just discover an interesting relationship between Borrower Rate and Current Credit Line, let's plot the scatter plot separately

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #plot current credit lines Vs borrower rate
ggplot(data=subset(df, !is.na(IncomeRange)), aes(y=BorrowerRate, x=CurrentCreditLines)) +
  geom_point( size=0.5, position='jitter')+
  scale_x_log10(breaks=seq(0,100,5))
```

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
#calculate pearson correlation between current credit lines and borrower rate
cor.test(df_s$BorrowerRate,df_s$CurrentCreditLines)
```
?cor.test

#Multivariate 

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #plot the Term of loan Vs loan original amount for each income range category in one plot
ggplot(data=subset(df, !is.na(IncomeRange)), aes(x=Term, y=LoanOriginalAmount,color=IncomeRange)) +
  geom_line(stat='summary',fun.y=mean)+
  scale_color_brewer(type='div')
```

It could be strange that not employed borrowers loan, in average more than borrows that earn more (0-24,999$ income range)
Borrowers whose income is more than 100,000 dollars turn to have the greatest loan original amount, it seems interesting to discover, why do they borrow this money, in other words, in what Listing category they are interested in.
We start by converting ,ListingCategory..numeric. to a categorical variable

df_s$ListingCategory..numeric. <- factor(df_s$ListingCategory..numeric.)

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #plot the borrower rate Vs loan original amount for each income range category in one plot
ggplot(data=subset(df, !is.na(IncomeRange)), aes(x=BorrowerRate, y=LoanOriginalAmount/Term, group=1, color=IncomeRange)) +
  geom_point()+
 scale_color_brewer(type='seq')
```

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
#create a summarising data set grouped by main categorical variables and highliting mean numerical variables
df_by_income <- df_s%>%
  group_by(IncomeRange, LoanStatus, CurrentCreditLines, ListingCategory)%>%
  summarise(mean_loan=mean(LoanOriginalAmount), median_loan=median(LoanOriginalAmount), mean_payments=mean(OnTimeProsperPayments),median_payments=median(OnTimeProsperPayments), mean_rate=mean(BorrowerRate))
```

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 #plot the current credit lines Vs mean of original amount loan for each income range
ggplot(data=subset(df_by_income, !is.na(IncomeRange)), aes(x=CurrentCreditLines, y=mean_loan, color=IncomeRange))+  geom_point(alpha=0.7, size=1.5, position='jitter')+
  scale_y_log10()+
 scale_color_brewer(type='div')
```

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
#calculate the pearson correlation between current credit lines and mean of original amount loan
cor.test(df_by_income$CurrentCreditLines,df_by_income$mean_loan)
```

Before scaling y axis to log10, the data seemed to be spread, now it looks there exists an exponential relation between current credit lines and mean original amount loan by income range, this is confirmed by the correlation coefficient: 0.287

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
  #plot the current credit lines Vs mean of original amount loan for each income range
ggplot(data=subset(df_by_income, !is.na(IncomeRange)), aes(x=CurrentCreditLines, y=median_loan, color=IncomeRange))+  geom_point(alpha=0.7, size=1.5, position='jitter')+
  scale_y_log10()+
 scale_color_brewer(type='div')
```

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
cor.test(df_by_income$CurrentCreditLines,df_by_income$median_loan)
```

The relation between current credit lines and median loan seems to stronger than the relation that relates it with mean loan since the coefficient is 0.318

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 
ggplot(data=subset(df_by_income, !is.na(IncomeRange)), aes(x=CurrentCreditLines, y=median_loan, color=LoanStatus))+  geom_point(alpha=0.7, size=1.5, position='jitter')+
  scale_x_log10()+
  facet_wrap(~ListingCategory)+
  geom_smooth(method='lm', color='red')
 scale_color_brewer(type='div')
```

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 
ggplot(data=subset(df_by_income, !is.na(IncomeRange)), aes(x=20*round(mean_payments/20), y=mean_loan))+  geom_line()+
  scale_x_continuous(breaks=seq(0,100,5))+
 scale_color_brewer(type='div')
```

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
cor.test(df_by_income$mean_payments,df_by_income$mean_loan)
```

As the correlation between the mean payment and mean loan is so weak (0.04), 

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
df_b_gg <- df_by_income %>%
  select(IncomeRange, LoanStatus, CurrentCreditLines, mean_loan, median_loan, mean_payments, median_payments, mean_rate)
```


```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 
library(GGally)
set.seed(1598)
theme_set(theme_minimal(20))
ggpairs(df_b_gg[sample.int(nrow(df_b_gg),1000),])
```

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
cor.test(df_by_income$CurrentCreditLines,df_by_income$mean_rate)
```

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
cor.test(df_by_income$mean_rate,df_by_income$mean_loan)
```



```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 
ggplot(data=subset(df_by_income, !is.na(IncomeRange)), aes(x=mean_loan, y=mean_rate, color=ListingCategory))+  geom_point(alpha=0.5, size=0.5)+
  scale_x_log10()+
 scale_color_brewer(type='div')
```

We notice a negative correlation between mean rate and mean loan especially for debt consolidation and home improvement, as the mean loan amount increases, the rate decreased, we could be interested in calculating by how much of dollars the loan increases.
>To do that we'll create a new variable, called increase:

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
df_s$increase <- df_s$LoanOriginalAmount-(df$LoanOriginalAmount*df$BorrowerRate)
```

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
df_by_income <- df_s%>%
  group_by(IncomeRange, LoanStatus, CurrentCreditLines, ListingCategory)%>%
  summarise(mean_loan=mean(LoanOriginalAmount), median_loan=median(LoanOriginalAmount), mean_payments=mean(OnTimeProsperPayments),median_payments=median(OnTimeProsperPayments), mean_rate=mean(BorrowerRate), mean_increase=mean(increase))
```


```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
cor.test(df_s$BorrowerRate,df_s$increase)
```

#Final Plots

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 
ggplot(data=df_s, aes(x=LoanOriginalAmount)) +
  geom_histogram( binwidth=500, color= I('black'), fill=I('#099DD9')) +
  xlab("loan original amount")+
  ggtitle("Distribution of loan original amounts")
  scale_x_continuous(breaks=seq(1000,28000,2000), limits=c(1000,30000))
```

Principally, there is 3 loan amounts that are more frequent: 4000, 10000 and 15000 dollars.

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 
ggplot(data=subset(df_s, !is.na(CurrentCreditLines)), aes(x=CurrentCreditLines, y=OnTimeProsperPayments)) +
  geom_jitter(alpha=0.7, size=1)+
xlab("current credit lines")+
  ylab("On time prosper payments")+
  ggtitle("Current credit lines Vs. On time prosper payments")
```

Expecting that it will be a negative correlation between current credit lines and on time prosper payments, it turns to be 0.06! the distribution seems to be somewhat normal and we also notice the presence of some outliers; for example, someone has about 60 credit lines at a time, but not even 50 on time payments!

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
 
ggplot(data=subset(df_by_income, !is.na(IncomeRange)), aes(x=CurrentCreditLines, y=mean_loan, color=IncomeRange))+  geom_point(alpha=0.7, size=1.5, position='jitter')+
  scale_y_log10()+
 scale_color_brewer(type='div')
```

```{r,echo=FALSE, results='asis', message = FALSE, error = FALSE, warning= FALSE}
cor.test(df_by_income$CurrentCreditLines,df_by_income$mean_loan)
```

Before scaling y axis to log10, the data seemed to be spread, now it looks there exists an exponential relation between current credit lines and mean original amount loan by income range, this is confirmed by the correlation coefficient: 0.287

#Reflections
I wanted to challenge myself by choosing a so complex data set, the first difficulty I encountered is the high number of features, I just read the variables description, and choose mentally most relevant ones, maybe by taking a course in feature selection I'll do better.
I think that I success in interpreting some key plots, discovering main trends and digest large amount of information.
